Combination of field and experimental data together with computational models reveal cognitive mechanisms behind cleaner fish behaviour

### 2022-02-16 manuscript_1.0/manuscript_1.0.tex
Fixed data files. Updated the cpp to g++ compilation. new parameter files to run chains with the fixed data files. Update figures in the manuscript

Disentangling the mechanistic underpinnings of behaviour
is typically done under controlled conditions in a laboratory setting. However,
controlled laboratory conditions often mask the inter and intra-specific
variation that arises in the interaction between mechanisms and
environmental variation. From a evolutionary perspective, mechanisms
are likely selected because of how they allow individuals to respond to
environmental variation. For example, biological market theory predicts that
the exchange rate of goods and/or services traded between cooperative
partners adjusts to the law of supply and demand, when individuals have
some degree of partner choice \citep{noe_Biological_1995a}.
Supply and demand conditions, which typically
depend on the abundance of the species involved, certainly vary in time
and space. Therefore, natural selection should
favour the ability to flexibly adjust decisions and behavioural output to
current market conditions. Indeed, such adjustments have been repeatedly
documented \citep{axen_Signalling_1996}. In animals, an obvious general candidate
mechanism for the strategic adjustment is the cognitive machinery. However,
it is not clear which cognitive mechanisms allow individuals to adjust their
behaviour to the varying market conditions. As adjustments to changes in
market conditions have been documented in a great variety of species with
highly variable brain anatomies, the question arises to what extent mechanisms
beyond basic associative learning may be involved.

@@

Recent research on the cognitive tool kit needed to solve the ephemeral
reward task has used a broad approach as proposed by Shettleworth
\citep{shettleworth_Cognition_2009}, who defined cognition as including
all ways in which animals acquire information
through the senses, process, retain and decide to act on it.
Indeed, perception of relevant cues is of major importance
for individual performance. For example, some animals find relevant
information in the plates
\citep{wismer_Cuebased_2019}) while others find it in the food
\citep{pretot_Comparative_2021, pretot_Comparing_2016}. However, identifying
cues as salient is not the only challenge of the ephemeral reward task,
as revealed by proximate learning models. Such models allow varying the
cognitive tool kit and to evaluate which minimal kit is necessary to solve
the task at hand (e.g. \citep{dubois_Model_2021}). Applied to the ephemeral
reward task, learning models showed that basic reinforcement learning does
not suffice to solve the ephemeral reward task \citep{prat_Modelling_2022, quinones_Reinforcement_2019}.
This is particularly so when models assume the more complex natural
situation in which cleaner fish not only face resident-visitor pairs but also
visitor-visitor and resident-resident pairs, as well a resident alone or
a visitor alone. To be able to give visitors priority over residents,
cleaners need to be able to assess a client's value separately for the
three possible scenarios (alone, paired with a fish with the
same strategic option, paired with a fish with the alternative strategic option)
\citep{quinones_Reinforcement_2019}. The ability to distinguish and value differently
one stimulus alone from compound versions of it has been termed
configurational learning, or chunking,
or segmentation (see references in \citep{prat_Modelling_2022}).

To solve the ephemeral reward task, cleaners need to account for
the future consequences of current decisions. In the model by
Quiñones \emph{et al.} \citep{quinones_Reinforcement_2019}, this could be achieved in
two non-mutually exclusive ways: through low temporal discounting of
future effects, also termed `chaining' \citep{enquist_Power_2016}; and/or
through perceiving a visitor client leaving as psychological punishment
(i.e.~as a negative reinforcer). Low temporal discounting is when
individuals include in their valuation of an action the reward effects
that this will have in the future. This is done by combining in a single
valuation the reward obtained in the current time with all the reward that
comes after, discounting for how far in the future the future reward
is accrued. `Chaining' the reward of these different time steps allows
individuals to take actions that increase the long-term reward at the sacrifice
of short term considerations \citep{enquist_Power_2016}. Even though, `chaining'
can be readily implemented computationally in learning models
\citep{enquist_Power_2016, sutton_Reinforcement_2018}, cognitively it seems to be
a complex adaptation \citep{suddendorf_Evolution_2007}. On the other hand,
using client behaviour as a negative reinforcer is in principle
easier to implement. Negative reinforcement is part of the ubiquitous
learning mechanism termed operant conditioning \citep{thorndike_Animal_1898, skinner_Behavior_1938}. Thus, standard logic of
Lloyd Morgan's cannon demands that operant conditioning as the simpler
explanation is to be accepted by default. Ideally, the two mechanisms should
be evaluated in light of how well they explain the available data.

Over the last decade, our team tested over a hundred of wild-caught
cleaner fish in the exact same paradigm of of the ephemeral reward
task \citep{salwiczek_Adult_2012, wismer_Variation_2014, triki_Decrease_2018, triki_Biological_2019, triki_Brain_2020}. These fish often come from
different reef locations around the study field site at Lizard Island,
Great Barrier Reef in Australia. Further investigation of the local
eco-sociological conditions revealed that cleaner and client
fish population densities have a substantial impact on cleaner fish
performance in the task. Cleaner fish from reef sites with relatively
low densities were more likely to fail solving the task \citep{triki_Biological_2019, triki_Decrease_2018, wismer_Variation_2014}. The explanation for this pattern
is twofold. First, low cleaner fish population densities imply
low supply and high demand for cleaning services.
Under such conditions, there are fewer occasions under
which failing to give priority to a visitor translates to an empty
cleaning station. Second and also related to the state of the cleaning market,
visitor clients at low cleaner fish population densities
are less likely to exert their partner choice options
and are hence more likely to wait for inspection if not given cleaning
priority \citep{triki_Brain_2020}. In the absence of partner choice
cleaners should not give priority to visitors any more.

The interplay between the underling cognitive machinery and
local ecological conditions apparently generate the documented variable
performance among individuals of the same species.
Our approach of fitting the computational model to the empirical data on
fish densities and cleaner fish censuses and cleaner fish performance
in the ephemeral reward task aimed at: (i), determining which
mechanism cleaner fish use to incorporate future consequences of current
decisions by testing whether operant conditioning, chaining, or a combination
of both best explains their performance; (ii) determining whether the
two mechanisms differed with respect to the ecological conditions
that are likely to cause high versus low performance in the
ephemeral reward task. Additionally, we assessed which mechanism
yields optimal performance patterns. Relying on the logic of
biological market theory, we predicted
that appropriate performance is to show low preference for visitors under low
local cleaner-to-client ratio. In this case, as visitor leaving rates
should be low as well and any leaving visitor will be readily
replaced by another client in the queue. In contrast,
cleaners should show high performance in the ephemeral reward task
if the local cleaner-to-client ratio favours clients.

@@

\hypertarget{overview}{%
\subsubsection{Overview}\label{overview}}

To test the relative validity of chaining versus negative reinforcement as
mechanisms to account for future consequences of current choices,
we combined empirical data from Triki \emph{et. al}
\citep{triki_Biological_2019, triki_Brain_2020}. with the computational
model analysed in Quiñones \emph{et al.}\citep{quinones_Reinforcement_2019}. In order to
compare the explanatory power of the different hypothesis, we estimate the
parameter values of three versions of the model that best fit the data.
Particularly, we fit the parameter that determines how important
future rewards are (\(\gamma\)), the magnitude of the negative reward
(\(\eta\)) and a scaling constant that translates absolute to relative
abundances (see following sections for details). We obtained model
predictions using a set of stochastic individual-based
simulations. Thus, each parameter combination of the model yields not one
prediction for the data set, but a distribution of them. Thus, for the
fitting procedure, we use one of the most common algorithms in bayesian inference,
the Markov Chain Monte Carlo (MCMC). MCMC is a stochastic algorithm that allows
to sample from the posterior distribution of model parameter values. The posterior
distribution represents the most likely parameter values given the data.
Due to the stochastic nature of MCMC, which repeatedly samples parameter values
and evaluates their likelihood, the distribution of likelihoods from the
stochastic model is also explored. This makes MCMC and bayesian inference
particularly suitable for the computational models we evaluate. Bellow,
we explain first the computational model and the data collection, and later
how the MCMC was used to estimate parameter values.

@@

The model consist of a set individual-based simulations where
individuals face a series of choices between two options. Individuals in
the model represent cleaners, and the options they face simulate the
natural conditions of a cleaning market. Cleaners experience a series
of discrete time points in which they face different `states', which
are defined by the number and type of clients inviting for service.
There are six possible states: zero clients, one resident, one visitor,
two residents, two visitors, and a mix of one resident and one visitor.
The probability of each state is largely determined by the relative
abundance of cleaners, residents and visitors, but to some degree by a
cleaner's choices when it faces the resident-visitor combination.
This is because residents are those clients that stay in the
cleaning queue when they are not given priority and hence would
still be available in the next time point; while visitors leave
the queue (with a certain probability) after they are not given
priority. Individuals obtain a fixed reward from cleaning a client
regardless of the type. Every time individuals face and make a choice
they update the probability of making that same choice. The update is
based on the difference between the expectation of reward and the actual
reward the individual obtains (prediction error)
\citep{sutton_Reinforcement_2018, rescorla_Theory_1972}; and it is carried
in the direction that would lead to more reward being obtained, given
the new information. In the long run, the probability of choosing a visitor
over a resident converges in the model. To which probability the model
will converge depends on the relative abundance of cleaners, visitors
and resident; as well as on the probability of visitors leaving the station
when unattended. Further details of the model implementation can be
found in the paper \citep{quinones_Reinforcement_2019}.

The model shows that only agents that calculate rewards separately for
the different states (termed `fully aware agents') can potentially
prioritise visitors of residents in nature, and hence also solve the
ephemeral reward task. This is because pooling rewards for the two
clients from all states by `partially aware agents'
(that only recognize the two clients as different) cancels out
all differences (see \citep{quinones_Reinforcement_2019}). However,
such \emph{chunking} by fully aware agents is not enough to develop a
preference for visitors. In addition, agents need to find a mean
to incorporate future consequences of current choices. In the model,
this could be achieved with either of two parameters that could
also work together. First, \(\gamma\) measures how
much individuals include future rewards in their decision updates. If
\(\gamma=0\), individuals only use the immediate reward obtained from a
cleaning interaction. As \(\gamma\) increases individuals include more the
reward obtained from the subsequent choices. That amounts to calculating
and using for decision making the future expected rewards of an action.
Second, \(\eta\) measures how much individuals include in their reward the
fleeing behaviour of visitor clients as a negative component. Both of
these parameters allow individuals to use in their estimates the future
effects of their choices.

A further model analysis contrast the effect of these two
parameters against variation in the relative abundance of cleaners
and the two client types. Assuming certainty that visitors leave, the
two parameters yield different
predictions when relative client abundance is very high. Under such
conditions, the analysis shows that under very high client (relative)
abundance the presence of future rewards (\(\gamma>0\)) no longer favours
the preference for the ephemeral clients, while the preference persists
when the negative reward mediated by client behaviour (\(\eta>0\)) is
included. Thus, the analysis suggests that future reward as an
explanation for cleaner ability to prefer visitors is supported.
At first sight, empirical data seem to support future rewards as
the mechanism used by cleaners, as the statistical analysis by
Triki \emph{et al.} \citep{triki_Biological_2019} data revealed that low
cleaner-to-client ratios predict low preference for visitors and hence
low performance in the ephemeral reward task. However, low cleaner-to-client
ratios may also lead to low probabilities of visitors leaving, which would
cause low preferences for visitors independently of the underlying
mechanism \citep{quinones_Reinforcement_2019}. Furthermore, a combination
of future and negative rewards, could still be yield a better explanation
for the data than any single factor.

@@

The empirical data were collected between 2010 and 2018 always during
the austral winter months June to August at Lizard Island
(\(14.6682° S, 145.4604° E\)), Great Barrier Reef, Australia.
They consist of three sets: fish densities from a total of five study
reef sites (Corner Beach-CB, Horseshoe-HS,
Mermaid Cove-MC, Northern Horseshoe-NHS, and The Crest-TC),
natural observations of cleaners to quantify the probability
of visitors leaving if initially ignored in favour of another client, and the
performance of individuals caught from these sites in the ephemeral reward test.
In total, we have twelve site/year data sets for fish densities and
corresponding performance in the test. Thus, some sites were sampled more
than once. Due to major environmental perturbations during the study period
(\(\color{red}{\text{refs}}\)), fish densities changed significantly even
within sites \citep{triki_Decrease_2018, triki_Fluctuations_2019}. To estimate
the population density of cleaner fish and their clients at a given site in
a given year, Triki \emph{et al.} \citep{triki_Biological_2019} used a series of ten
transects of \(30m\) each. Observers swam along the transect lines placed on
the reef and first counted the visible large-bodied adult fish
(species with total length TL \textgreater{} \(10cm\)) on a width of \(5m\), and then on the
return individuals of small-bodied fish species (TL \textless{} \(10 cm\)) on a width
of \(1 m\). We included cleaner fish in the \(5 m\) transect. Based on the
classification by Bshary \citep{bshary_Cleaner_2001a} and Wismer \emph{et al.}
\citep{wismer_Variation_2014}, all visitor client species (\(n = 124\)) were
large-bodied. Resident clients comprised all small-bodied species
(\(n = 30\)) but also some large-bodied ones (\(n = 15\)). Overall,
large-bodied species are highly correlated with being a visitor,
and small-bodied species are a good correlate of residents. Based on the
transect data, Triki \emph{et al.} \citep{triki_Biological_2019, triki_Brain_2020}
estimated cleaner fish, small-bodied client, large-bodied client and total
client fish densities per \(100 m^2\). LASSO analyses revealed that cleaner and
large-bodied client densities were highly correlated, and only the former
was hence kept in the analyses as representative of both \citep{triki_Biological_2019}.
Cleaner densities, in interaction with visitor client leaving probability,
explained best individual performance in the ephemeral reward task
\citep{triki_Biological_2019}, while cleaner-to-client ratio was not a significant
factor. This creates a certain mismatch with the model by Quiñones \emph{et al.}
\citep{quinones_Reinforcement_2019}, which focused on abundance of cleaners relative
to overall client abundance in order to capture the characteristics
of the market. Importantly, relative abundances of clients define not only
the probability of interactions with residents and visitors, but also affect
how often the cleaning station is empty (e.g.~there are no clients to be cleaned).
The higher the relative abundance of cleaners is, the higher is the probability
of cleaning stations being empty at a given time point. How often the
cleaning stations are empty in turns influences the value that cleaners can
expect from their choices. The emptier the cleaning stations are, the more
beneficial it is to prioritize a client that will not stay. On the other hand,
situations where cleaners have to choose will be rare if cleaner relative
abundance is high, which reduces the total value of prioritizing visitors
over residents and makes it more difficult to learn a preference for visitors.
Given the model structure, we used the empirical data to calculate relative
abundances of cleaners, residents and visitors for the current analyses.
To incorporate the empirical finding that absolute densities matter, we
introduced a new scaling constant (see section on model fitting).

To collect data on the visitor clients' probability of leaving if not serviced
immediately, eight cleaners per site/year were each filmed for \(30 min\).
From the videos, all events in which a visitor client was made to wait in
favour of another (visitor or resident) client were identified, and noted
whether or not the visitor left \citep{triki_Biological_2019, triki_Brain_2020}.
A total of 120 cleaners (10 individuals per 12 site/year) were tested in the
ephemeral reward task \citep{triki_Biological_2019, triki_Brain_2020}. Authors
housed all captured cleaners individually in glass aquaria
( \(62cm \times 27cm \times 37 cm\) ) and provided them
with PVC pipes (\(10 cm \times 1 cm\)) as shelters.
The task consisted of exposing the cleaner fish to substitute
models of client fish in the form of two \emph{Plexiglas} plates offering the
same amount of food (one item of mashed prawn). The two plates differed
in colour and pattern (horizontal green stripes or vertical pink stripes)
but had equal size (\(10 cm \times 7 cm\)). Importantly, the two plates played
different roles as either a visitor (ephemeral food source) or
resident (permanent food source). That is, if a cleaner inspected the
resident plate first, the experimenter withdrew the visitor plate out of
the aquarium as a consequence. Choosing first the visitor plate,
however, granted access to both plates. The equal size of the plates
forced cleaner fish to learn to give service priority to the visitor plate
based solely on the behaviour-cue of the plates rather than size-cue
\citep{wismer_Cuebased_2019}. Triki \emph{et al.} \citep{triki_Biological_2019, triki_Brain_2020}
tested the fish for a maximum of 200 trials with 20 trials a day, 10 trials
in the morning and 10 trials in the afternoon. They randomized and
counterbalanced the plates' spatial location (i.e.~left or right)
between trials. Similarly, they counterbalanced the plates' decoration
(colour and pattern) and the plates' role (visitor or resident) between the
tested fish. The data was originally collected to assess which variables are
good predictors of the capacity to solve the ephemeral food task. Thus, previous
analysis use specific statistical criteria to classify individual cleaners'
performance as successful or failure \citep{triki_Biological_2019}. Because the model
predicts the preference of cleaners given the set of environmental variable,
we instead use the raw data of individuals cleaners' decisions.
We use 20 choices from each cleaner as the preference they develop
for the visitor plate. Which these 20 choices were vary for each cleaner depending
on the performance that was originally assessed. For cleaners that did not
succeed, the 20 choices correspond to the last two sessions. Cleaners that
succeeded in solving the task were tested on a reversal task. That is, the
pattern in plates was switched to the opposite role they played in the
initial test. For those cleaners, we used the last session before they succeeded
in the initial, and the last session of the reversal, completing the 20 choices.
We did not distinguish between individuals that reached the criterion early
or late during the 200 trials.

@@

The aim of the analysis is to fit the key model parameters that allow
individuals to incorporate future consequences of current actions,
\(\gamma\) and \(\eta\), to the data from Triki \emph{et al.}
\citep{triki_Biological_2019, triki_Brain_2020} to find out whether each or
a combination of these effects is a better explanation for the pattern
seen in the data. In particular, we are interested in the relation between the
ecological variables: cleaners, visitor clients, resident clients abundances and
visitor clients leaving probability; and cleaners' preference for visitor clients
over residents. Thus, the four ecological variables are given as input to
the models. The preference for visitors gotten from the simulations of the model
is used as the prediction for the response variable, while the number of visitors
chosen out of the 20 choices of each cleaner as the observed response variable.
We kept all other parameter values used for the model simulations constant,
see Table \ref{tab:param}.

In order to capture with the model the relationship between the characteristics
of the market (environment) and the cleaner's preference we need to translate
the absolute abundances of the data to to a measure of relative abundances
that captures client visitation patterns. This is because in the model, relative
abundances of clients define not only the probability of residents and visitors,
but also how often the cleaning station is empty (e.g.~there are no
clients to be cleaned). The frequency with which clients visit the station
is another variable, which will likely vary among different client species
depending on parasite abundances and loads. We do not have field estimates
for species specific parasite loads, especially not as a function of site.
In order to control for these aspects, we computed a measure of relative
cleaner abundance for each site relative to absolute abundances and multiplied
by a scaling constant that changes the range of the variable. This scaling
constant is hence meant to capture variation in the market conditions driven
partly by cleaner abundance. We fit the value of the scaling constant as
part of the statistical inference. As for the client abundances for
residents and visitors we computed a relative measure with respect to the
total client abundance, and weighted that by the rescaled relative absence
of cleaners (\(1-\text{relative cleaner abundance}\)). Thus, all three measures
of relative abundance sum up to 1, and can be used in the model as a proxy
for the probability of having different options in the cleaning station.
Note that the scaling constant was introduced to control for variation that
is not captured by the model; its parameter distribution does not offer
biological insights.

>>>>

Disentangling the mechanistic underpinnings of behaviour
is typically done under controlled conditions in a laboratory setting. However,
controlled laboratory conditions often mask the inter and intra-specific
variation that arises in the interaction between mechanisms and
environmental variation. From a evolutionary perspective, mechanisms
are likely selected because of how they allow individuals to respond to
environmental variation. For example, biological market theory predicts that
the exchange rate of goods and/or services traded between cooperative
partners adjusts to the law of supply and demand, when individuals have
some degree of partner choice \citep{noe_Biological_1995a}.
Supply and demand conditions, which typically
depend on the abundance of the species involved, certainly vary in time
and space. Therefore, natural selection should
favour the ability to flexibly adjust decisions and behavioural output to
current market conditions. Indeed, such adjustments have been repeatedly
documented \citep{axen_Signalling_1996}. In animals, an obvious general candidate
mechanism for the strategic adjustment is the cognitive machinery. However,
it is not clear which cognitive mechanisms allow individuals to adjust their
behaviour to the varying conditions. As adjustments to changes in
market conditions have been documented in a great variety of species with
highly variable brain anatomies, the question arises to what extent mechanisms
beyond basic associative learning may be involved.

@@

Recent research on the cognitive tool kit needed to solve the ephemeral
reward task has used a broad approach as proposed by Shettleworth
\citep{shettleworth_Cognition_2009}, who defined cognition as including
all ways in which animals acquire information
through the senses, process, retain and decide to act on it.
Indeed, perception of relevant cues is of major importance
for individual performance. For example, some animals find relevant
information in the plates
\citep{wismer_Cuebased_2019} while others find it in the food
\citep{pretot_Comparative_2021, pretot_Comparing_2016}. However, identifying
cues as salient is not the only challenge of the ephemeral reward task,
as revealed by proximate learning models. Such models allow varying the
cognitive tool kit and to evaluate which minimal kit is necessary to solve
the task at hand (e.g. \citep{dubois_Model_2021}). Applied to the ephemeral
reward task, learning models showed that basic reinforcement learning does
not suffice to solve the ephemeral reward task \citep{prat_Modelling_2022, quinones_Reinforcement_2019}.
This is particularly so when models assume the more complex natural
situation in which cleaner fish not only face resident-visitor pairs but also
visitor-visitor and resident-resident pairs, as well a resident alone or
a visitor alone. To be able to give visitors priority over residents,
cleaners need to be able to assess a client's value separately for the
three possible scenarios (alone, paired with a fish with the
same strategic option, paired with a fish with the alternative strategic option)
\citep{quinones_Reinforcement_2019}. The ability to distinguish and value differently
one stimulus alone from compound versions of it has been termed
configurational learning, or chunking,
or segmentation (see references in \citep{prat_Modelling_2022}).

To solve the ephemeral reward task, cleaners need to account for
the future consequences of current decisions. In the model by
Quiñones \emph{et al.} \citep{quinones_Reinforcement_2019}, this could be achieved in
two non-mutually exclusive ways: through low temporal discounting of
future effects, also termed `chaining' \citep{enquist_Power_2016}; and/or
through perceiving a visitor client leaving as psychological punishment
(i.e.~as a negative reinforcer). Low temporal discounting is when
individuals include in their valuation of an action the reward effects
that this will have in the future. This is done by combining in a single
valuation the reward obtained in the current time with all the reward that
comes after, discounting for how far in the future reward
is accrued. `Chaining' the reward of these different time steps allows
individuals to take actions that increase the long-term reward at the sacrifice
of short term considerations \citep{enquist_Power_2016}. Even though, `chaining'
can be readily implemented computationally in learning models
\citep{enquist_Power_2016, sutton_Reinforcement_2018}, cognitively it seems to be
a complex adaptation \citep{suddendorf_Evolution_2007}. On the other hand,
using client behaviour as a negative reinforcer is in principle
easier to implement. Negative reinforcement is part of the ubiquitous
learning mechanism termed operant conditioning \citep{thorndike_Animal_1898, skinner_Behavior_1938}. Thus, standard logic of
Lloyd Morgan's cannon demands that operant conditioning as the simpler
explanation is to be accepted by default. Ideally, the two mechanisms should
be evaluated in light of how well they explain the available data.

Over the last decade, our team tested over a hundred of wild-caught
cleaner fish in the exact same paradigm of of the ephemeral reward
task \citep{salwiczek_Adult_2012, wismer_Variation_2014, triki_Decrease_2018, triki_Biological_2019, triki_Brain_2020}. These fish often come from
different reef locations around the study field site at Lizard Island,
Great Barrier Reef in Australia. Further investigation of the local
eco-sociological conditions revealed that cleaner and client
fish population densities have a substantial impact on cleaner fish
performance in the task. Cleaner fish from reef sites with relatively
low densities were more likely to fail at solving the task \citep{triki_Biological_2019, triki_Decrease_2018, wismer_Variation_2014}. The explanation for this pattern
is twofold. First, low cleaner fish population densities imply
low supply and high demand for cleaning services.
Under such conditions, there are fewer occasions under
which failing to give priority to a visitor translates to an empty
cleaning station. Second and also related to the state of the cleaning market,
visitor clients at low cleaner fish population densities
are less likely to exert their partner choice options
and are hence more likely to wait for inspection if not given cleaning
priority \citep{triki_Brain_2020}. In the absence of partner choice
cleaners should not give priority to visitors any more.

The interplay between the underling cognitive machinery and
local ecological conditions apparently generate the documented variable
performance among individuals of the same species.
Our approach of fitting the computational model to the empirical data on
fish densities and cleaner fish censuses and performance
in the ephemeral reward task aimed at: (i), determining which
mechanism cleaner fish use to incorporate future consequences of current
decisions by testing whether operant conditioning, chaining, or a combination
of both best explains their performance; (ii) determining whether the
two mechanisms differed with respect to the ecological conditions
that are likely to cause high versus low performance in the
ephemeral reward task. Additionally, we assessed which mechanism
yields optimal performance patterns. Relying on the logic of
biological market theory, we predicted
that appropriate performance is to show low preference for visitors under low
local cleaner-to-client ratio.

@@

@@

The model consists of a set individual-based simulations where
individuals, representing cleaners, face a series of choices
between two options, which simulate the natural conditions
of the cleaning market. Individuals experience a series
of discrete time points in which they face different `states',
defined by the number and category of client fish (visitor or residents)
inviting for cleaning services. There are six possible states:
zero clients, one resident, one visitor,
resident-resident, visitor-visitor, and resident-visitor.
The probability of each state is largely determined by the relative
abundance of cleaner fish, residents and visitors, but to some degree by
cleaner fish choices when it faces the resident-visitor combination.
This is because residents are willing to queue for cleaning service;
while visitors leave the queue (with a certain probability) when made to wait.
Individuals obtain a fixed reward from cleaning a client fish
regardless of the category. Every time individuals face and make a choice
they update the probability of making that same choice. The update is
based on the difference between the expected reward and the obtained
reward (prediction error) \citep{sutton_Reinforcement_2018, rescorla_Theory_1972}. Given the new information, the update is carried
in the direction that leads to more reward being obtained, given
the new information. In the long run, the probability of choosing a visitor
over a resident converges in the model. To which probability the model
will converge depends on the relative abundance of cleaners, visitors
and resident; as well as on the probability of visitors leaving the
cleaning station when made to wait. Further details of the model
implementation can be found in Quiñones \emph{et al} \citep{quinones_Reinforcement_2019}.

The model shows that agents need to find a mean
to incorporate future consequences of current choices. In the model,
this could be achieved with either of two parameters that could
also work together. First, \(\gamma\) measures how
much individuals include future rewards in their decision updates. If
\(\gamma=0\), individuals only use the immediate reward obtained from a
cleaning interaction. As \(\gamma\) increases individuals include more the
reward obtained from the subsequent choices. That amounts to calculating
and using for decision making the future expected rewards of an action.
Second, \(\eta\) measures how much individuals include in their reward the
fleeing behaviour of visitor as a negative component. Both of
these parameters allow individuals to use in their estimates the future
effects of their choices.

@@

The empirical data were collected between 2010 and 2019 always during
the austral winter months June to August from a total of five
study reef sites (Corner Beach-CB, Horseshoe-HS, Mermaid Cove-MC,
Northern Horseshoe-NHS, and The Crest-TC) at Lizard Island
(\(14.6682° S, 145.4604° E\)), Great Barrier Reef, Australia.
The data consist of three sets: fish censuses, field
observations of cleaner-client interactions to quantify the probability
of visitors leaving if made to wait, and the
performance of wild-caught cleaner fish in the ephemeral reward test.
In total, we have twelve site/year data sets for fish censuses and
corresponding performance in the lab test. Thus, some sites were sampled more
than once. To estimate the population density of cleaner fish
and their clients at a given site in
a given year, Triki \emph{et al.} \citep{triki_Biological_2019} used a series of ten
transects of \(30m\) each. Observers swam along the transect lines placed on
the reef and first counted the visible large-bodied adult fish
(species with total length \(TL \geq 10cm\)) including cleaner fish
on a width of \(5m\), and then on the
return individuals of small-bodied fish species (\(TL < 10 cm\))\\
on a width of \(1 m\)
(see Triki et al.~2018 for further details on fish censuses data collection).
We then scaled the counts of cleaner fish, small-bodied,
and large-bodied clients fish densities per \(100 m^2\).
From the study by Triki \emph{et al} \citep{triki_Biological_2019}
cleaner fish and large-bodied client populations densities were
highly correlated, and only the former
was hence used in the analyses as representative of both measures.

The field observation data consisted of video recordings/encodings of the
cleaner-client cleaning interactions. There were videos from eight
cleaners per site/year of a duration of \(30 min\) each. Triki \emph{et al}
extracted information from every event wherein a visitor client
was made to wait in favour of another client (visitor or resident), and noted
whether or not the visitor left or queued for the cleaning service
\citep{triki_Biological_2019, triki_Brain_2020}.

The cognitive performance data was from a total of 120 cleaners
(10 individuals per 12 site/year) tested in the
ephemeral reward task \citep{triki_Biological_2019, triki_Brain_2020}. Authors
housed all captured cleaners individually in glass aquaria
( \(62cm \times 27cm \times 37 cm\) ) and provided them
with PVC pipes (\(10 cm \times 1 cm\)) as shelters.
The task consisted of exposing the cleaner fish to substitute
models of client fish in the form of two \emph{Plexiglas} plates offering the
same amount of food (one item of mashed prawn). The two plates differed
in colour and pattern (horizontal green stripes or vertical pink stripes)
but had equal size (\(10 cm \times 7 cm\)). Importantly, the two plates played
different roles as either a visitor (ephemeral food source) or
resident (permanent food source). That is, if a cleaner inspected the
resident plate first, the experimenter withdrew the visitor plate out of
the aquarium as a consequence. Choosing first the visitor plate,
however, granted access to both plates. The equal size of the plates
forced cleaner fish to learn to give service priority to the visitor plate
based solely on the behaviour-cue of the plates rather than size-cue
\citep{wismer_Cuebased_2019}. Triki \emph{et al.} \citep{triki_Biological_2019, triki_Brain_2020}
tested the fish for a maximum of 200 trials with 20 trials a day, 10 trials
in the morning and 10 trials in the afternoon. They randomized and
counterbalanced the plates' spatial location (i.e.~left or right)
between trials. Similarly, they counterbalanced the plates' decoration
(colour and pattern) and the plates' role (visitor or resident) between the
tested fish. In the original studies, once a fish reach a learning criterion,
that is, performing significantly above chance level (\(> 50%
\), \(p-value ≥ 0.05\)),
they passed to a reversal version of the task where the roles of the
visitor/resident Plexiglas plates were swapped (see Triki et al.~2019, 2020).
Here, we used instead a subset of these data in order to have an idea on
cleaner fish preferences for the visitor plate, even if they do not reach
the learning criterion within 200 trials. To do so, we first extracted the
trial-by-trial outcomes from the last two session (20 trials) of those who
never reached the learning criterion for visitor plate
\(\color{red}{\text{(N = XX cleaner fish)}}\).
For those who reached the learning criterion at some point during the
test and passed to a reversal phase, we extracted the trials-by-trial
outcomes from the last session (10 trials) before passing to reversal
and the last 10 trials they were exposed to in the test
\(\color{red}{\text{(N = XX cleaner fish)}}\).

@@

The aim of the analyses is to fit the key model parameters
\(\gamma\) and \(\eta\), to the empirical data from Triki \emph{et al.}
\citep{triki_Biological_2019, triki_Brain_2020} to test whether each or
a combination of these effects is a better explanation for the pattern
seen in the data. We used the
ecological variables: cleaners, visitor clients,
resident clients abundances and
visitor clients leaving probability, as input to the models.
As response variable we used cleaners' preference for visitor
over residents in the ephemeral reward task.
Finally, we used the preference for visitors resulting from the
model simulations as the prediction for the response variable.
We kept all other parameter values used for the model simulations constant,
see Table \ref{tab:param}.

To capture with the model the relationship between the ecological
variables and cleaner fish preferences for visitors we need to scale
the absolute population densities of cleaner fish from the empirical data
to a measure of relative abundances that captures
client visitation patterns. This is because in the model, relative
abundances of clients define not only the probability of residents and visitors,
but also how often the cleaning station is empty (e.g.~there are no
clients to be cleaned). The frequency with which clients visit the station
is another variable, which in nature may vary among different client species
depending on their ectoparasite loads. We do not have field estimates
for species-specific parasite loads, especially not as a function of site.
In order to control for these aspects, we computed a measure of relative
cleaner fish abundance for each reef site relative to
absolute abundances and multiplied it
by a scaling constant that changes the range of the variable. This scaling
constant is hence meant to capture variation in the market conditions driven
partly by cleaner abundance. We fitted the value of the scaling constant as
part of the statistical inference. As for the visitor and resident
abundances we computed a relative measure with respect to the
total client abundance, and weighted that by the rescaled relative absence
of cleaners (\(1-\text{relative cleaner abundance}\)). Thus, all three measures
of relative abundance sum up to 1, and can be used in the model as a proxy
for the probability of having different options in the cleaning station.
Note that we introduced the scaling constant to control for variation that
is not captured by the model; its parameter distribution does not offer
biological insights.

Once we calculated the relative abundances, we obtained predictions from
the model for each one of the locations and ran the Markov Chain. We started the
chain with random values for the parameters of interest, then ran the
computational model once for every reef site using as input the ecological
explanatory variables. The model outputs the probability of choosing the visitor
for each location \(p_i\), where \(i\) is the index for the 12 locations. Assuming a
binomial distribution, the probability that each of the cleaner fish 20 choices
in the ephemeral reward task was generated by the model
is given by \(\binom{n}{k_ij}p^{k_{ij}}_i (1-p_i)^{n-k_{ij}}\);
where \(k_{ij}\) is the number of times that cleaner fish \(j\) in reef site \(i\)
chose the visitor over the resident; and \(n=20\) (due to our choice of using 20
choices per cleaner fish). By taking the natural logarithm and summing over all
the cleaner fish and reef sites we obtained the log-likelihood of the data
given the model and parameters. We then proposed a new set of parameters drawn
from a uniform distribution centred around the old parameter set. The amplitude
of the uniform distribution used for each parameter can be found in Table
\ref{tab:param}. Subsequently, we ran the model and calculated the likelihood
with the new parameter set. We then used the ratio of the two likelihoods to
choose which parameter set to keep. New parameter sets with a higher likelihood
than the old set replaced old ones, and those with a lower likelihood replaced
current ones with a probability equals to the log-likelihood ratio.
Given that we only used the likelihoods in the decision, we used a
uninformative prior. Once we decided whether the new parameter set replaces
the old one, we ran the model again to sample the likelihood
distribution of the parameter set. We then started
the cycle again by proposing a new set of parameters, and repeated
the process for \(1e^5\) steps. We ran 5 independent chains,
discarded the first \(100\) samples of each chain as burn in,
and after that we kept 1 in every 100 samples to avoid autocorrelation.
The collection of parameters sets kept in
all chains approximates the posterior distribution. We coded the model
as well as the fitting algorithm in \emph{c++}, the diagnostics and visualization
in R \citep{rcoreteam_Language_2021}. All codes are accessible at
\(\color{red}{\text{include citation}}\)

To compare the fit of the three alternative models, we used the
distribution of pseudo \(R^2\) proposed by Mcfadden \citep{mcfadden_Conditional_1974}.
Mcfadden's pseudo \(R^2\) is a standard measure of fit for logistic regression.
In that context, \(pseudo-R^2\) uses the log-likelihood of the data given
the model, relative to the log-likelihood of the data given a
model without covariates, as a measure of fit. Our
model is not a logistic regression, therefore we measured the \(pseudo-R^2\) in
relation to the log-likelihood of a model
with parameters \(\gamma\) and \(\eta\) set to zero. This in practice
amounts to a model that has neutral preference between the two options.
We computed the pseudo \(R^2\) for all the samples of the posterior from
the MCMC. Thus, we used these distributions of pseudo \(R^2\)'s as a measure of
fit.
